Equalizer\\Programming and User Guide

### 2009-11-18 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.1 for Equalizer 0.9+svn\\[\medskipamount]
%  \today
}

@@

Written by Stefan Eilemann.\\
Engineering contributions by Maxim Makhinya, Jonas B\"osch, Christian
Marten and Patrick Bouchaud.

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{Subpixel Compounds}

Subpixel decomposes the rendering into different jittered frusta so 
that the recomposition anti-aliases the final rendering.
The Subpixel compound is also compatible with applications doing their own software 
anti-aliasing.

@@

Subpixel compounds in Equalizer are configured using the \textsf{subpixel}
parameter, using the values \textsf{[ index size ]}. The \textsf{index} 
parameter corresponds to the current ressource's index and the \textsf{size} 
is the total number of ressources needed.

The \textsf{index} has to be smaller than the ressources \textsf{size}. 
\fig{fSubpixel} illustrates a four-way subpixel decomposition.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/subpixel.pdf}
  {\caption{\label{fSubpixel}A four-way Subpixel compound}}
\end{figure}

@@

\subsubsection{Writing Compounds}
The following steps are typically taken when writing compound
configurations:
\begin{description}
\item [Root compound] Define an empty top-level compound when
  synchronizing multiple destination views. Multiple destination views
  are used for multi-display systems, e.g., a PowerWall or CAVE. All
  windows used for one display surface should be swap-locked (see below)
  to provide a seamless image. A single destination view is typically
  used for providing scalability to a single workstation window.
\item [Destination compound(s)] Define one compound for each destination
  channel, either as a child of the empty group, or as a top-level
  compound.
  \begin{description}
  \item[Standalone channels] Set the channel by using the channel name
    from the resource section. Define a frustum as explained in
    \sref{sCanvas}.
  \item[View/Segment channels] Set the channel by using the canvas,
    segment, layout and view name. The compound frustum will be
    calculated automatically based on the segment or view frustum. Note
    that one segment may created multiple view/segment channels, one for
    each view intersection of each layout used on the canvas. Only the
    compounds belonging to the active layout of a canvas are activate at
    runtime.
    \begin{description}
    \item [Software swap lock] When using software swap synchronization,
      swap-lock the destination compounds using a swapbarrier. All
      windows with a swapbarrier of the same name synchronize their
      swapbuffers.
    \item [Hardware swap lock] When using hardware swap synchronization,
      use swapbarriers for the destination compounds, setting
      \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
      swap barrier name is ignored in this case. All windows of the same
      group on a single node synchronize their swap buffer. All groups
      of the same barrier synchronize their swap buffer across
      nodes. Please note that the driver typically limits the number of
      groups and barriers to one, and that multiple windows per swap
      group are not supported by all drivers. Please refer to your
      OpenGL driver documentation for details.
    \end{description}
  \end{description}
\item[Scalability] If desired, define scalability for each of your
  destination compounds. Add one compound using a source channel for
  each contributor to the rendering. The destination channel may also be
  used as a source. 
  \begin{description}
  \item[Decomposition] On each child compound, limit the rendering task
    of that child by setting the \textsf{viewport}, \textsf{range},
    \textsf{period} and \textsf{phase}, \textsf{pixel}, \textsf{eye} or
    \textsf{zoom} as desired.
  \item[Runtime Adjustments] A \textsf{load\_equalizer} may be used on
    the destination compounds to set the \textsf{viewport} or
    \textsf{range} of all children each frame, based on the current
    load. A \textsf{view\_equalizer} may be used on the root compound to
    assign resources to all destination compounds, which have to use
    \textsf{load\_equalizer}s. A \textsf{framerate\_equalizer} should be
    used to smoothen the framerate of DPlex compounds. A
    \textsf{DFR\_equalizer} may be used to set the zoom of a compound to
    achieve a constant framerate. One compound may have multiple
    equalizers, e.g., a \textsf{load\_equalizer} and a
    \textsf{DFR\_equalizer} for a 2D compound with a constant framerate.
  \item[Recomposition] For each source compound, define an
    \textsf{output\_frame} to read back the result. Use this output frame
    as an \textsf{input\_frame} on the destination compound. The frames
    are connected with each other by their name, which has to be unique
    within the root compound tree. For parallel compositing, describe
    your algorithm by defining multiple input and output frames across
    all source compounds.
  \end{description}
\end{description}

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination channel's rendering, which is defined by the viewport,
frustum and database. A \textsf{viewport} decomposes the destination
channel and frustum in screen space. A \textsf{range} tells the
application to render a part of its database, and an \textsf{eye}
rendering pass can selectively render different stereo passes. A
\textsf{pixel} parameter adjusts the frustum so that the source channel
renders an even subset of the parent's pixels. A \textsf{subpixel} parameter
tells the source channels to render different jittered frustum to perform
anti-aliasing. Setting one or multiple attributes causes the parent's view
to be decomposed accordingly. Attributes are cumulative, that is,
intermediate compound attributes affect and therefore decompose the
rendering of all their children.

@@

\paragraph{Load Equalizer}
\vspace{-2ex}\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D Load-Balancing}}
\end{wrapfigure}
While pixel and stereo compounds are naturally load-balanced, 2D and DB
compounds often need load-balancing for optimal rendering
performance. Using a load equalizer is transparent to the application,
and can be used on any application for 2D, and on most applications for
DB load-balancing. Some applications do not support dynamic updates of
the database range, and therefore cannot be used with DB load-balancing.

@@

\paragraph{View Equalizer}
Depending on the model position and data structure, each segment of a
multi-display system has a different rendering load. The segment with
the biggest load determines the overall performance when using a static
assignment of resources to segments. The view equalizer analyzes the
load of all segments, and adjusts the resource usage each frame. It
equalizes the load on all segments of a view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
      Load-Balancing for two Segments using eight GPU's}}
\end{wrapfigure}
The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particularly useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

\begin{wrapfigure}{l}{.382\textwidth} 
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
        Load-Balancing for a CAVE}}
\end{wrapfigure}
\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPU's. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPU's from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

@@

\paragraph{DFR Equalizer}
Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adjusts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and ray-tracing.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}Dynamic Frame Resolution}}
\end{wrapfigure}
\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

@@

\part{Programming Guide}

@@

The particulars of creating a window and OpenGL context depend on the
window system used. One can either use the implementation provided by
the operating system, e.g., AGL, WGL or glX, or some higher-level
toolkit, e.g., Qt.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}
\end{wrapfigure}
All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

@@

\begin{description}
\item[Buffer] The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  separation and anaglyphic stereo settings.
\item[Viewport] The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum] The same frustum parameters as defined by
  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
  projection matrix. The frustum is influenced by the destination
  channel's view definition, compound viewports, head matrix and the
  current eye pass. If the anti-aliasing idle mode is set, the frustum
  will be jittered by a vector. This part is described in \ref{sFSAA}.
\item[Head Transformation] A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frusta in immersive rendering. It is normally used to set up the
  `view' part of the modelview matrix, before static light sources are
  defined.
\item[Range] A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data. It is influenced by the
  compound range attribute.
\end{description}

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[h!t]\center
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\label{fRender}Main Render Loop}}
\end{figure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visibility against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

\textsf{frameAssemble} is overwritten for the use of Subpixel compound.
The whole process is described in \ref{sFSAA}.

@@

\begin{wrapfigure}{r}{.382\textwidth}
\vspace{-2ex}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}
\vspace{-4ex}
\end{wrapfigure}
Events are received by an event handler. The event handler finds the
\textsf{eq::OSWindow} associated to the event. It then creates a generic
\textsf{Event}, which holds the event data in an independent
format. The original native event and this generic \textsf{Event}
form the \textsf{OSWindowEvent}, which is passed to the concrete
\textsf{OSWindow} for processing.

The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

@@

Equalizer provides an \textsf{AGLEvent\-Handler},
\textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
events for an \textsf{AGLWindowIF}, \textsf{GLXWindowIF} and
\textsf{WGLWindowIF}, respectively. \fig{fEventUML} illustrates the
class hierarchy for event processing.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

The application or node main thread is the primary thread of each
process and executes the \textsf{main} function. The application and
render clients initialize the local node for communications with other
nodes, including the server, using \textsf{Client::initLocal}.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one Node Process}}
\end{wrapfigure}
During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame}
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2ex}
  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
  \vspace{-5ex}
\end{wrapfigure}
Changing view frustum parameters is typically done for non-fullscreen
application windows and multi-view layouts, where the rendering is not
meant to be viewed in real-world size in an immersive environment. A
typical use case is changing the field-of-view of the rendering.

A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
application process can set the wall or projection parameters at
runtime. For a description of wall and projection parameters please refer
to \sref{sFrustum}. The new data will be effective for the next frame.

@@

To provide a motivation for the design of the image compositing
API, the direct send parallel compositing algorithm is introduced in this
section. Other parallel compositing algorithms, e.g. binary-swap, can
also be expressed through an Equalizer configuration file.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Direct Send Compositing}}
\end{wrapfigure}
The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

@@

The frame data is a holder for images and additional information, such
as output frame attributes and pixel data availability.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
\end{wrapfigure}
The frame synchronization through the frame data allows the input frame
to wait for the pixel data to become ready, which is signaled by the
output frame after readback.

@@

When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

\subsection{\label{sFSAA}Full-scene and Software Anti-Aliasing}

The eqPly example contains support for software anti-aliasing to show 
how to combine it with the Equalizer Subpixel compound.

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.2 for Equalizer 0.9+svn\\[\medskipamount]
%  \today
}

@@

Written by Stefan Eilemann.\\
Engineering contributions by Maxim Makhinya, Jonas B\"osch, Christian
Marten, Sarah Amsellem and Patrick Bouchaud.

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{\label{sSubpixel}Subpixel Compounds}

Subpixel decomposes the rendering of multiple samples for one pixel, e.g., for
anti-aliasing or depth-of-field rendering. The default implementation provides
transparent, scalable software idle-antialiasing when configuring a subpixel
compound.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
  {\caption{\label{fSubpixel}\small A Subpixel Compound}}
\end{wrapfigure}
Applications can use subpixel compounds to accelerate depth-of-field effects and
software anti-aliasing, with potentially a different number of idle or non-idle
samples per pixel.

@@

Subpixel compounds are configured using the \textsf{subpixel} parameter, using
the values \textsf{[ index size ]}. The \textsf{index} parameter corresponds to
the current resource index and the \textsf{size} is the total number of
resources used.

The \textsf{index} has to be smaller than the \textsf{size}. 
\fig{fSubpixel} illustrates a three-way subpixel decomposition.

@@

@@

\subsubsection{Writing Compounds}
The following steps are typically taken when writing compound
configurations:
\begin{description}
\item [Root compound] Define an empty top-level compound when
  synchronizing multiple destination views. Multiple destination views
  are used for multi-display systems, e.g., a PowerWall or CAVE. All
  windows used for one display surface should be swap-locked (see below)
  to provide a seamless image. A single destination view is typically
  used for providing scalability to a single workstation window.
\item [Destination compound(s)] Define one compound for each destination
  channel, either as a child of the empty group, or as a top-level
  compound.
  \begin{description}
  \item[Standalone channels] Set the channel by using the channel name
    from the resource section. Define a frustum as explained in
    \sref{sCanvas}.
  \item[View/Segment channels] Set the channel by using the canvas,
    segment, layout and view name. The compound frustum will be
    calculated automatically based on the segment or view frustum. Note
    that one segment may created multiple view/segment channels, one for
    each view intersection of each layout used on the canvas. Only the
    compounds belonging to the active layout of a canvas are activate at
    runtime.
    \begin{description}
    \item [Software swap lock] When using software swap synchronization,
      swap-lock the destination compounds using a swapbarrier. All
      windows with a swapbarrier of the same name synchronize their
      swapbuffers.
    \item [Hardware swap lock] When using hardware swap synchronization,
      use swapbarriers for the destination compounds, setting
      \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
      swap barrier name is ignored in this case. All windows of the same
      group on a single node synchronize their swap buffer. All groups
      of the same barrier synchronize their swap buffer across
      nodes. Please note that the driver typically limits the number of
      groups and barriers to one, and that multiple windows per swap
      group are not supported by all drivers. Please refer to your
      OpenGL driver documentation for details.
    \end{description}
  \end{description}
\item[Scalability] If desired, define scalability for each of your
  destination compounds. Add one compound using a source channel for
  each contributor to the rendering. The destination channel may also be
  used as a source. 
  \begin{description}
  \item[Decomposition] On each child compound, limit the rendering task of that
    child by setting the \textsf{viewport}, \textsf{range}, \textsf{period} and
    \textsf{phase}, \textsf{pixel}, \textsf{subpixel}, \textsf{eye} or
    \textsf{zoom} as desired.
  \item[Runtime Adjustments] A \textsf{load\_equalizer} may be used on
    the destination compounds to set the \textsf{viewport} or
    \textsf{range} of all children each frame, based on the current
    load. A \textsf{view\_equalizer} may be used on the root compound to
    assign resources to all destination compounds, which have to use
    \textsf{load\_equalizer}s. A \textsf{framerate\_equalizer} should be
    used to smoothen the framerate of DPlex compounds. A
    \textsf{DFR\_equalizer} may be used to set the zoom of a compound to
    achieve a constant framerate. One compound may have multiple
    equalizers, e.g., a \textsf{load\_equalizer} and a
    \textsf{DFR\_equalizer} for a 2D compound with a constant framerate.
  \item[Recomposition] For each source compound, define an
    \textsf{output\_frame} to read back the result. Use this output frame
    as an \textsf{input\_frame} on the destination compound. The frames
    are connected with each other by their name, which has to be unique
    within the root compound tree. For parallel compositing, describe
    your algorithm by defining multiple input and output frames across
    all source compounds.
  \end{description}
\end{description}

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the destination
channel's rendering, which is defined by the viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in screen
space. A \textsf{range} tells the application to render a part of its database,
and an \textsf{eye} rendering pass can selectively render different stereo
passes. A \textsf{pixel} parameter adjusts the frustum so that the source
channel renders an even subset of the parent's pixels. A \textsf{subpixel}
parameter tells the source channels to render different samples for one pixel to
perform anti-aliasing or depth-of-field rendering. Setting one or multiple
attributes causes the parent's view to be decomposed accordingly. Attributes are
cumulative, that is, intermediate compound attributes affect and therefore
decompose the rendering of all their children.

@@

\paragraph{Load Equalizer}
\vspace{-2ex}\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D Load-Balancing}}
\end{wrapfigure}
While pixel, subpixel and stereo compounds are naturally load-balanced, 2D and
DB compounds often need load-balancing for optimal rendering performance. Using
a load equalizer is transparent to the application, and can be used on any
application for 2D, and on most applications for DB load-balancing. Some
applications do not support dynamic updates of the database range, and therefore
cannot be used with DB load-balancing.

@@

\paragraph{View Equalizer}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
      Load-Balancing for two Segments using eight GPU's}}
\end{wrapfigure}
Depending on the model position and data structure, each segment of a
multi-display system has a different rendering load. The segment with
the biggest load determines the overall performance when using a static
assignment of resources to segments. The view equalizer analyzes the
load of all segments, and adjusts the resource usage each frame. It
equalizes the load on all segments of a view.

@@

\begin{wrapfigure}{r}{.382\textwidth} 
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
        Load-Balancing for a CAVE}}
\end{wrapfigure}
The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particularly useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPU's. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPU's from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

@@

\paragraph{DFR Equalizer}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}Dynamic Frame Resolution}}
\end{wrapfigure}
Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adjusts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and ray-tracing.

\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

@@

\clearpage
\part{Programming Guide}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}\vspace{-4ex}
\end{wrapfigure}
The particulars of creating a window and OpenGL context depend on the
window system used. One can either use the implementation provided by
the operating system, e.g., AGL, WGL or glX, or some higher-level
toolkit, e.g., Qt.

All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

@@

\begin{description}
\item[Buffer] The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  separation and anaglyphic stereo settings.
\item[Viewport] The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum] The same frustum parameters as defined by
  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL projection
  matrix. The frustum is influenced by the destination channel's view
  definition, compound viewports, head matrix and the current eye pass. If the
  channel has a subpixel parameter, the frustum will be jittered before it is
  applied. Please refer to \ref{sFSAA} for more information.
\item[Head Transformation] A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frusta in immersive rendering. It is normally used to set up the
  `view' part of the modelview matrix, before static light sources are
  defined.
\item[Range] A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data. It is influenced by the
  compound range attribute.
\end{description}

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[h!t]\center
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\label{fRender}Main Render Loop}}
\end{wrapfigure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visibility against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

The \textsf{frameAssemble} method is overwritten for the use of Subpixel
compound with idle software anti-aliasing, as described in \ref{sFSAA}.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::OSWindow} associated to the event. It then creates a generic
\textsf{Event}, which holds the event data in an independent
format. The original native event and this generic \textsf{Event}
form the \textsf{OSWindowEvent}, which is passed to the concrete
\textsf{OSWindow} for processing.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
Equalizer provides an \textsf{AGLEvent\-Handler},
\textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
events for an \textsf{AGLWindowIF}, \textsf{GLXWindowIF} and
\textsf{WGL\-Win\-dowIF}, respectively. \fig{fEventUML} illustrates the
class hierarchy for event processing.

The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one Node Process}}
\end{wrapfigure}
The application or node main thread is the primary thread of each
process and executes the \textsf{main} function. The application and
render clients initialize the local node for communications with other
nodes, including the server, using \textsf{Client::initLocal}.

During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame}
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

Changing view frustum parameters is typically done for non-fullscreen
application windows and multi-view layouts, where the rendering is not
meant to be viewed in real-world size in an immersive environment. A
typical use case is changing the field-of-view of the rendering.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
\end{wrapfigure}
A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
application process can set the wall or projection parameters at
runtime. For a description of wall and projection parameters please refer
to \sref{sFrustum}. The new data will be effective for the next frame.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Direct Send Compositing}}
  \vspace{-3ex}
\end{wrapfigure}
To provide a motivation for the design of the image compositing
API, the direct send parallel compositing algorithm is introduced in this
section. Other parallel compositing algorithms, e.g. binary-swap, can
also be expressed through an Equalizer configuration file.

The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
\end{wrapfigure}
The frame data is a holder for images and additional information, such
as output frame attributes and pixel data availability.

@@

The frame synchronization through the frame data allows the input frame
to wait for the pixel data to become ready, which is signaled by the
output frame after readback.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

\subsection{\label{sFSAA}Custom Subpixel Processing}

The eqPly example contains support for software anti-aliasing when being idle,
showing how to combine it with subpixel compounds.

\subsubsection{Supersampling Algorithm}
